The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-05-22 Manuscript/manuscript.qmd
Re-add missing Procedure section; fix appendix link

The experimental task was programmed in JavaScript, using packages from the Phaser physics engine (https://phaser.io) and the jsPsych library [@deleeuwJsPsychJavaScriptLibrary2015]. The stimuli, presented on a black background, consisted of a circular blue ball -- controlled by the participant via the mouse or trackpad cursor; a rectangular green target; a red rectangular barrier located between the ball and the target; and an orange square within which the participant could control the ball before releasing it in a throw towards the target. Because the task was administered online, the absolute distance between stimuli could vary depending on the size of the computer monitor being used, but the relative distance between the stimuli was held constant. Likewise, the distance between the center of the target and the training and testing locations was scaled such that relative distances were preserved regardless of screen size. For the sake of brevity, subsequent mentions of this relative distance between stimuli, or the position where the ball landed in relation to the center of the target, will be referred to simply as distance. @fig-IGAS_Methods displays the layout of the task, as it would appear to a participant at the start of a trial, with the ball appearing in the center of the orange square. Using a mouse or trackpad, participants click down on the ball to take control of the ball, connecting the movement of the ball to the movement of the cursor. Participants can then "wind up" the ball by dragging it (within the confines of the orange square) and then launch the ball by releasing the cursor. If the ball does not land on the target, participants are presented with feedback in red text at the top right of the screen, on how many units away they were from the center of the target. If the ball was thrown outside of the boundary of the screen participants are given feedback as to how far away from the target center the ball would have been if it had continued its trajectory. If the ball strikes the barrier (from the side or by landing on top), feedback is presented telling participants to avoid hitting the barrier. If participants drag the ball outside of the orange square before releasing it, the trial terminates, and they are reminded to release the ball within the orange square. If the ball lands on the target, feedback is presented in green text, confirming that the target was hit, and presenting additional feedback on how many units away the ball was from the exact center of the target.

[Link to abbrevaited example of task](https://pcl.sitehost.iu.edu/tg/demos/igas_expt1_demo.html){target="_blank"}.

@@

@@

To prepare the data, we first removed trials that were not easily interpretable as performance indicators in our task. Removed trials included: 1) those in which participants dragged the ball outside of the orange starting box without releasing it, 2) trials in which participants clicked on the ball, and then immediately released it, causing the ball to drop straight down, 3) outlier trials in which the ball was thrown more than 2.5 standard deviations further than the average throw (calculated separately for each throwing position), and 4) trials in which the ball struck the barrier. The primary measure of performance used in all analyses was the absolute distance away from the center of the target. The absolute distance was calculated on every trial, and then averaged within each subject to yield a single performance score, for each position. A consistent pattern across training and testing phases in both experiments was for participants to perform worse from throwing positions further away from the target -- a pattern which we refer to as the difficulty of the positions. However, there were no interactions between throwing position and training conditions, allowing us to collapse across positions in cases where contrasts for specific positions were not of interest. All data processing and statistical analyses were performed in R version 4.32 [@rcoreteamLanguageEnvironmentStatistical2020]. ANOVAs for group comparisons were performed using the rstatix package [@kassambaraRstatixPipeFriendlyFramework2021a].

@@

[Apppendix](Appendix/Full_Appendix.qmd) available at tegorman13.github.io/Dissertation/Sections/Appendix.html

>>>>

The experimental task was programmed in JavaScript, using packages from the Phaser physics engine (https://phaser.io) and the jsPsych library [@deleeuwJsPsychJavaScriptLibrary2015]. The stimuli, presented on a black background, consisted of a circular blue ball - controlled by the participant via the mouse or trackpad cursor; a rectangular green target; a red rectangular barrier located between the ball and the target; and an orange square within which the participant could control the ball before releasing it in a throw towards the target. Because the task was administered online, the absolute distance between stimuli could vary depending on the size of the computer monitor being used, but the relative distance between the stimuli was held constant. Likewise, the distance between the center of the target and the training and testing locations was scaled such that relative distances were preserved regardless of screen size. For the sake of brevity, subsequent mentions of this relative distance between stimuli, or the position where the ball landed in relation to the center of the target, will be referred to simply as distance. @fig-IGAS_Methods displays the layout of the task, as it would appear to a participant at the start of a trial, with the ball appearing in the center of the orange square. Using a mouse or trackpad, participants click down on the ball to take control of the ball, connecting the movement of the ball to the movement of the cursor. Participants can then "wind up" the ball by dragging it (within the confines of the orange square) and then launch the ball by releasing the cursor. If the ball does not land on the target, participants are presented with feedback in red text at the top right of the screen, specifying how many scaled units away the ball was from the center of the target. If the ball was thrown outside of the boundary of the screen participants are given feedback as to how far away from the target center the ball would have been if it had continued its trajectory. If the ball strikes the barrier (from the side or by landing on top), feedback is presented telling participants to avoid hitting the barrier. If participants drag the ball outside of the orange square before releasing it, the trial terminates, and they are reminded to release the ball within the orange square. If the ball lands on the target, feedback is presented in green text, confirming that the target was hit, and presenting additional feedback on how many units away the ball was from the exact center of the target.

[Link to abbreviated example of task](https://pcl.sitehost.iu.edu/tg/demos/igas_expt1_demo.html){target="_blank"}.

@@

#### Procedure

Participants first electronically consented to participate, and then read instructions for the task which explained how to control the ball, and the goal of throwing the ball as close to the center of the target as possible. The training phase was split into 10 blocks of 20 trials, for a total of 200 training trials. Participants in the constant condition trained exclusively from a single location (760 scaled units from the target center). Participants in the varied condition trained from two locations (610 and 910 scaled units from the target center), encountering each location 100 times. The sequence of throwing locations was pseudo-random for the varied group, with the constraint that within every block of 20 training throws both training locations would occur 10 times. Participants in both conditions also received intermittent testing trials after every 20 training trials. Intermittent testing trials provided no feedback of any kind. The ball would disappear from view as soon as it left the orange square, and participants were prompted to start the next trial without receiving any information about the accuracy of the throw. Each intermittent testing stage consisted of two trials from each of the three training positions (i.e. all participants executed two trials each from Positions 610, 760, and 910 during each of the 10 intermittent testing stages). Following training, all participants completed a final testing phase from four positions: 1) their training location, 2) the training location(s) of the other group, 3) a location novel to both groups. The testing phase consisted of 15 trials from each of the four locations, presented in a randomized order. All trials in the final testing phase included feedback. After finishing the final testing portion of the study, participants were queried as to whether they completed the study using a mouse, a trackpad,  or some other device (this information was used in the exclusion process described above). Finally, participants were debriefed as to the hypotheses and manipulation of the study.

@@

To prepare the data, we removed trials that were not easily interpretable as performance indicators in our task. Removed trials included: 1) those in which participants dragged the ball outside of the orange starting box without releasing it, 2) trials in which participants clicked on the ball, and then immediately released it, causing the ball to drop straight down, 3) outlier trials in which the ball was thrown more than 2.5 standard deviations further than the average throw (calculated separately for each throwing position), and 4) trials in which the ball struck the barrier. The primary measure of performance used in all analyses was the absolute distance away from the center of the target. The absolute distance was calculated on every trial, and then averaged within each subject to yield a single performance score, for each position. A consistent pattern across training and testing phases in both experiments was for participants to perform worse from throwing positions further away from the target -- a pattern which we refer to as the difficulty of the positions. However, there were no interactions between throwing position and training conditions, allowing us to collapse across positions in cases where contrasts for specific positions were not of interest. All data processing and statistical analyses were performed in R version 4.32 [@rcoreteamLanguageEnvironmentStatistical2020]. ANOVAs for group comparisons were performed using the rstatix package [@kassambaraRstatixPipeFriendlyFramework2021a].

@@

[Apppendix](https://tegorman13.github.io/Dissertation/Sections/Appendix/Full_Appendix.html) available at https://tegorman13.github.io/Dissertation/Sections/Appendix/Full_Appendix.html

### 2024-05-22 Presentation/slides.qmd
Re-add missing Procedure section; fix appendix link

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
subtitle: "Dissertation Defense"
format: 
  html: 
    toc: true
    #shift-heading-level-by: 1
    #toc-expand: true
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    output-file: slides_revealjs
    slideNumber: true
    center: true
    chalkboard: true
  pptx:
    incremental: true
author:
  - name: Thomas Gorman
    url: https://tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations:
      - name: Indiana University
        department: Psychological & Brain Sciences, Cognitive Science
        address: "55 Park Place NE, #464"
        city: Bloomington
        region: IN
        country: USA
        postal-code: 47401
---

::::: {.content-visible when-format="html" unless-format="revealjs"}
\
\
\
\
\
\
\
\
\
\

# Slides {{< fa photo-film size=2xs >}} {.unnumbered .unlisted visibility="hidden"}

@@

:::: column-page-inset-right
::: {.ratio .ratio-16x9}
<iframe src="/Presentation/slides_revealjs.html">

</iframe>
:::
::::
:::::

# Introduction

@@

# Project 1 - Hit The Target (HTT)

@@

# Project 2 - Hit The Wall (HTW)

@@

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e1
//| fig-cap: "Experiment 1 Design. Constant and Varied participants complete different training conditions."
//| fig-width: 7.0
//| fig-height: 2.5
//| fig-responsive: false
//| echo: false  
///| eval: !expr out_type 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]

  // define the global styles of the nodes
  node [shape = rectangle, style = filled]

  data1 [label = " Varied Training \n800-1000\n1000-1200\n1200-1400", fillcolor = "#FF0000"]
  data2 [label = " Constant Training \n800-1000", fillcolor = "#00A08A"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]

  // edge definitions with the node IDs
  data1 -> Test1
  data2 -> Test1

  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }

  Test2 -> Test3
}
```
:::

@@

# General Discussion

@@

## Comparison {.smaller}

| Dimension | HTT (Project 1) | HTW (Project 2) |
|----------------|--------------------------------|------------------------|
| Task Description | Projectile launching to hit a target | Projectile launching to hit wall at a specific velocity |
| Task Complexity | More complex parabolic trajectory, both x and y velocities relevant to outcome | Simpler 1D mapping of force to outcome. Only x velocity is relevant. |
| Task Space | More complex: xy velocity combinations closer to the solution space may still result in worse feedback due to striking the barrier. | Simpler: smooth, linear mapping between velocity and feedback. |
| Perceptual salience of Varied Conditions | Varied conditions (# of throwing distances) are perceptually distinct, i.e. salient differences in distance between launching box and target. | Varied conditions (# of velocity bands) are less salient - only difference is the numeral displayed on screen. |
| Testing Feedback | Testing always included feedback | Primary testing stage had no feedback. |
| Potential for Learning during Testing | Limited potential for learning during testing due to feedback. | Some potential for learning during no-feedback testing by observing ball trajectory. |
| Training Experience | Varied group gets half as much experience on any one position as the constant group. | Varied group gets 1/3 as much experience on any one velocity band as the constant group. |
| Testing Structure | Random interleaving of trained/transfer testing distances. | Blocked structure, separately testing trained vs extrapolation testing bands. |

@@

>>>>

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
subtitle: "Dissertation Defense"
format: 
  html: default
    #toc: true
    #shift-heading-level-by: 1
    #toc-expand: true
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    output-file: slides_revealjs
    slideNumber: true
    center: true
    chalkboard:
      theme: whiteboard
  #   scrollable: true
  # pptx:
  #   incremental: true
  # hugo-md:
  #   include: false
  #   html-math-method: mathjax
  #   output-file: slides_hugo.md
  gfm:
    output: false
    output-file: slides_gfm.md
author:
  - name: Thomas Gorman
    url: https://tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations:
      - name: Indiana University
        department: Psychological & Brain Sciences, Cognitive Science
---

::: {.content-visible when-format="html" unless-format="revealjs"}
\
[Click here](../paper.qmd) to open the dissertation manuscript.\
\
\
\
\
\
\
\
\
\

## Slides {{< fa photo-film size=2xs >}} {.unnumbered .unlisted visibility="hidden"}

@@

::: column-body-outset-right
::: {.ratio .ratio-16x9}
<iframe src="/Presentation/slides_revealjs.html">

</iframe>
:::
:::

\
\
\
\
:::

## Introduction

@@

## Project 1 - Hit The Target (HTT)

@@

## Project 2 - Hit The Wall (HTW)

@@

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e1
//| fig-cap: "Experiment 1 Design. Constant and Varied participants complete different training conditions."
//| fig-width: 7.0
//| fig-height: 2.5
//| echo: false  
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n800-1000\n1000-1200\n1200-1400", fillcolor = "#FF0000"]
  data2 [label = " Constant Training \n800-1000", fillcolor = "#00A08A"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```
:::

@@

## General Discussion

@@

## Comparison {.smaller .r-fit-text}

| Dimension                                | HTT (Project 1)                                                                                                                               | HTW (Project 2)                                                                                                |
|------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------|
| Task Description                         | Projectile launching to hit a target                                                                                                          | Projectile launching to hit wall at a specific velocity                                                        |
| Task Complexity                          | More complex parabolic trajectory, both x and y velocities relevant to outcome                                                                | Simpler 1D mapping of force to outcome. Only x velocity is relevant.                                           |
| Task Space                               | More complex: xy velocity combinations closer to the solution space may still result in worse feedback due to striking the barrier.           | Simpler: smooth, linear mapping between velocity and feedback.                                                 |
| Perceptual salience of Varied Conditions | Varied conditions (# of throwing distances) are perceptually distinct, i.e. salient differences in distance between launching box and target. | Varied conditions (# of velocity bands) are less salient - only difference is the numeral displayed on screen. |
| Testing Feedback                         | Testing always included feedback                                                                                                              | Primary testing stage had no feedback.                                                                         |
| Potential for Learning during Testing    | Limited potential for learning during testing due to feedback.                                                                                | Some potential for learning during no-feedback testing by observing ball trajectory.                           |
| Training Experience                      | Varied group gets half as much experience on any one position as the constant group.                                                          | Varied group gets 1/3 as much experience on any one velocity band as the constant group.                       |
| Testing Structure                        | Random interleaving of trained/transfer testing distances.                                                                                    | Blocked structure, separately testing trained vs extrapolation testing bands.                                  |

@@

<!-- create slide with 2 columns and add text -->

## Future Research Directions {.smaller .r-fit-text}

::: columns
::: {.column width="50%"}
### Theoretical Extensions

-   Investigate the role of individual differences in training variability effects.
-   Explore the impact of different types of feedback on learning and generalization.
-   Examine the influence of task complexity on the benefits of varied training.

### Methodological Innovations

-   Develop new experimental paradigms to test training variability in more complex tasks.
-   Utilize advanced computational models to simulate learning processes under varied conditions.
-   Implement longitudinal studies to assess long-term effects of training variability.
:::

::: {.column width="50%"}
### Practical Applications

-   Design adaptive training programs that adjust variability based on learner performance.
-   Apply findings to educational settings to enhance learning outcomes.
-   Use insights to improve training protocols in professional skill development.

### Collaborative Opportunities

-   Partner with educational institutions to test training variability in classroom settings.
-   Collaborate with industry professionals to apply research findings in real-world training programs.
-   Engage with interdisciplinary teams to explore the broader implications of training variability.
:::
:::
